---
title: "Replicating Breznau et al. 2022"
author: "Alexander Dean"
format:
  pdf:
    link-citations: true
    number-sections: true
    fontsize: 11pt
    indent: true
    keep-tex: true
    math: true
    documentclass: article
    number-depth: 3
    include-in-header: preamble.tex
execute:
  echo: false
  warning: false
  cache: true
---

# Introduction

\noindent Below details how I chose to replicate Breznau and coauthors' 2022 PNAS article, where they test the hypothesis that there is an association between immigration rates and support for government service delivery spending in sectors like old-age pensions, healthcare, etc. My version of replicating their results are not necessarily the most "right" way to replicate, but rather it is based on how I understood their paper. This exercise is intended to show how difficult replication is whenever you have unclear instructions. Thus, you may have chosen different coding choices. I've split my R code into specific sections. I've also chosen "echo: true" and "eval: true" so that you can see both the code that I used and the final output.

# Setup

\noindent This section details the equivalent of my YAML for the .r script. Note that it follows the best practices from the recitation slides for this week.

```{r}
#| eval: true
#| echo: true
#| warning: false

################################################################################
################# This R Script Does the Following: ############################
### Replicates the main analysis from Breznau et al. 2022 article in PNAS.
### Specifically, it tests the association between immigration rates and
### support for government service delivery cross-nationally.
################################################################################
### Set working directory, load packages, and load data
# Working Directory

# NOTE: set the working to YOUR correct pathway
setwd("/Users/alexdean/Documents/Vanderbilt/Service/Spring 2026/replication_exercise")
options(scipen = 999)

# Re-run if you are installing a package for the first time
if (!require(tidyverse)) install.packages("tidyverse")
if (!require(fixest)) install.packages("fixest")
if (!require(modelsummary)) install.packages("modelsummary")

# Or run this after: pacman::p_load(tidyverse, stargazer)

# Load data (CHANGE TO RELEVANT DIRECTORY PATHWAY FOR YOU):
# Individual-level data
survey <- haven::read_dta("data/individ_data.dta")
# Country-level data
macro <- read.csv("data/breznau_etal_2022_macro.csv")
```

# Wrangling and Cleaning

\noindent This section of code provides "pre-cleaning" of the data that will save you time as you do the replication exercise today. You can ignore this unless you'd like to see how I merged the relevant data and other pre-processing details.

``` {r}
#| echo: true
#| eval: true
#| warning: false

##### Alex Cleaning ############################################################
### Pre-cleaning survey df
# I'm removing the columns that start with c_[insert] that feature the government
# service delivery policy support questions on an expanded scale. Students will
# only have access to d_[insert] questions which are a binary 0 = no, 1 = yes
# Ignore this part if you are comparing your results to mine. This is for the 
# individual-level survey data. I am also creating a merge variable so I can 
# add macro-level country-year variables like immigration rate and immigration
# stock. Also also standardizing country strings

# Removing c_[insert] variables, age_squared variable, weighting variable, etc. 
# Also adding a new country_year variable for later merging.
survey <- survey |> 
  mutate(
    # Standardizing country strings
    country = country |> 
      str_to_lower() |> 
      str_remove_all("\\s+"),
    # Creating a new country_year variable for merging purposes
    country_year = paste(country, year, sep = "-"),
  ) |> 
  select(
    # Retaining only the columns I want
    country, year, country_year, starts_with("d_"), age, yrsedu
    )

### Pre-cleaning macro df
# I'm removing columns that are not relevant, renaming columns as needed, and 
# creating a new country_year variable for merging purposes
macro <- macro |> 
  mutate(
    # Standardizing country strings
    country = country |> 
      str_to_lower() |> 
      str_remove_all("\\s+"),
    # Creating a new country_year variable for merging purposes
    country_year = paste(country, year, sep = "-")
  ) |> 
  select(
    # Note migstock is migration stock variable from different agencies like 
    # the UN, Worldbank, etc. al_ethnic is the Alesina ethnic fractionalization
    # variable from his 2004 study
    country_year, migstock_wb, migstock_un, migstock_oecd, al_ethnic
  )

### Merging with macro-level variables at the country-year level
data <- left_join(
  survey, macro, by = "country_year"
)

##### Saving to .csv ###########################################################
write_csv(data, "replication_data.csv")
################################################################################
```

# Replication Process

\noindent This section starts the official replication process that you did today in recitation and details my wrangling and cleaning decision-making before I run the regression. Note my thoroughness in commenting my code. Any code comment that is preceded by "Alex:" details my logic as I was cleaning here. "Alex Cleaning" details cleaning that I did to make the df that you received generally legible and straightforward to replicate yourselves.

```{r}
#| eval: true
#| echo: true
#| warning: false

##### Cleaning #################################################################
### Missing Data
# Some columns in this dataframe are missing data. How will you handle units
# that do have some missing data? Be thorough in checking ALL relevant columns

# Alex: I am not removing individuals who have missing DV data. I can argue 
# that this data is essentially missing at random and that because we are making
# an index variable, we can reasonably assume that any missing answer wouldn't
# significantly change the indexed response. Of course, you may argue the 
# opposite! This shows the importance of being clear in replication. I am
# rows where no country was recorded since I plan on using country FE. I am also
# standardizing NAs for the migration stock variables using the R NA rather than
# the STATA NA.

data <- data |>
  filter(!str_starts(country_year, "-")) |>
  mutate(
    across(starts_with("migstock_"), ~ na_if(.x, "."))
  )

### Your DV
# Are you creating an index of government service delivery policies? Or are you 
# using one of these variables? 

# Alex: I am creating an index across each of these variables. By creating an
# index across conceptually different types of government service delivery 
# categories, I can argue that this increases our confidence that we are
# understanding the general effect of immigration rates on support for funding
# for government social policy. Of course, you may argue that it is better to 
# know if immigration stock rates have a different association with different
# types of government service, if, for example, immigrants are more likely to
# use government-funded healthcare than state housing stipends.
data <- data |> 
  mutate(
    # Creating the DV index
    dv_index = rowMeans(across(starts_with("d_")), na.rm = TRUE),
    # Changing any actual NAs from NaN to NA
    dv_index = na_if(dv_index, NaN)
    )

### Checking correlation of migstock variables
# None of the migration stock variables from the various agencies feature full
# coverage for the entire panel. Here I am checking the correlation between 
# migstock_un and migstock_wb because they seem to be fairly similar and
# migstock_wb doesn't have 2016 data while migstock_un does. If they
# are highly correlated, then I will create a single migstock variable that uses
# migstock_wb when available and migstock_un when the former is NA (in 2016).

# Creating separate df with only non-NAs for the two columns
overlap <- data |> 
  filter(!is.na(migstock_wb), !is.na(migstock_un)) |> 
  mutate(
    migstock_wb = as.numeric(migstock_wb),
    migstock_un = as.numeric(migstock_un)
  )

# Running a correlation between the two
cor_check <- cor(overlap$migstock_wb, overlap$migstock_un, use = "complete.obs")

# The correlation is .99! So we can reasonably do the plan from above
print(cor_check)

# Making one single migstock variable from migstock_un and migstock_wb and
# keeping only relevant variables
data <- data |> 
  mutate(
    migstock = as.numeric(coalesce(migstock_wb, migstock_un))
  ) |> 
  select(
    -c(migstock_wb, migstock_un, migstock_oecd)
  )
```

# Running the Regression

```{r}
#| echo: true
#| eval: true
#| warning: false
#| results: asis

##### Analysis #################################################################
### Regress immigration rate on support for xxx. Consider the following:
# 1) Are you including any control(s)?
# 2) Are you maintaining the panel or collapsing it?
# 3) Are you including fixed effects?
# 4) Are you clustering your std. errors? At what level?

reg <- feols(
  dv_index ~ migstock | country + year,
  data = data,
  cluster ~ country
)

# In your regression table, you should have the following:
# 1) A clearly printed table title
# 2) Clearly printed column label(s) / model name(s)
# 3) Clearly printed covariate labels (don't keep the variable name)
# 4) In the bottom of the table, print the number of observations, Adj. R2,
# if you used FEs. Do not include information like the BIC
# 5) In the table notes, include any information you believe to be relevant to 
# interpreting the table. Also include the legend for your p-value stars

reg_table <- modelsummary(
  models   = list("DV Index" = reg),
  title    = "Assc. Between Immigration and Govt. Policy Support",
  coef_map = c("migstock" = "Migration stock"),
  statistic = "({std.error})",
  stars = TRUE,

  gof_map = data.frame(
    raw   = c("nobs", "adj.r.squared"),
    clean = c("Observations", "Adjusted R2"),
    fmt   = c(0, 3)
  ),

  add_rows = data.frame(
    term = c("Country FE", "Year FE"),
    `DV Index` = c("Yes", "Yes")
  ),

  notes = c(
    "Note: Standard errors clustered by country.
    Significance levels: * p < 0.10, ** p < 0.05, *** p < 0.01."
  ),
  notes_append = TRUE,
  output = "latex" 
)

reg_table
```

\noindent Well, as we can see from the outcome, we fail to reject the null hypothesis that there is an association between migration stock rates and support for government social spending related to old age pensions, healthcare, etc. Despite literally finding a zero-point estimate, note that my results are substantively different in comparison to many of the teams that worked on this replication exercise. How do your results compare?
